Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-06-23 Manuscript/01-Intro.Rmd
More introduction writting

Many applications of modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see @Mallick2013. These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients: @Ravikumar2009 extended the grouped lasso [@Yuan2006] to additive models; @Huang2010 further developed adaptive grouped lasso for additive models; @Wang2007 and @Xue2009 applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed where the application on AMs are by-products. @Scheipl2012 proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma, specifically for semi-parametric regression using MCMC. The normal-mixture-of-inverse gamma prior is mixture of t-distribution with an additional coefficients mixing vector, where the vector follows a mixture of two Gaussian distribution. @Bai2020 was the first to apply group spike-and-slab lasso prior to AMs using a fast optimization algorithm, and further generalized the framework to GAMs in @Bai2021. Regardless of the nice statistical properties, a practical and pivotal question, if non-linear effects exist, yet to be addressed by these methods. In addition, a common criticism received by these methods is that the basis function coefficients tend to be overly penalized due to the sparsity assumption directly applied on the basis function coefficients.[@Scheipl2013] <!-- It is still not clear what is the best practice on how to model non-linear effects in high-dimensional setting.  -->

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data, where the smooth functions are imposed with a new spike-and-slab spline prior. Two fast computing optimization based algorithms are developed for high and ultra-high dimension settings respectively. The proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using mixture double exponential distribution. To avoid the over-penalization of the basis function coefficients, smoothness penalties are incorporated in the model via re-parameterization. Smoothness penalties are commonly implemented in GAMs through smoothing and penalized regression splines. They are quadratic norm of the coefficients, and allow locally adaptive penalties on each smoothing function. For the most common case of cubic splines, the smoothness penalty conditioning on a smoothness penalty parameter $\lambda_j$ is a function of the integration of second derivative of the spline function,  expressed mathematically as
$$
  \text{pen}f_j(x) = \lambda_j \int f^{"}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j,
$$
where $\bs S_j$ is a known penalty matrix, and $\bs \beta_j$ are the basis function coefficients. Smoothing penalties are also previously used in spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothness penalty allows the separation of the linear space of the smooth function, also know as the null space, from the nonlinear space. [TODO: descrive wood.]

Describe the differences from previous models SS GAM and Bai, taling about the computational burden.

Moreover, compared to currently available Bayesian spline models which mostly conduct a 'all-in-all-out' strategy, i.e. either including a non-linear form of a variable or excluding the variable completely, we offer a more flexible solution, rending one of three possibilities: no effect of a predictor, only linear effect of a predictor, or non-linear effect.

@@

Lastly, The third aim of the dissertation work will focus on the software implementation of the proposed models and algorithms. We will implement the proposed models and algorithms in one of the most commonly used statistical computing environment R. Ancillary functions will be provided in assistance to model diagnostic, interpretation and graphing. The R implementation of the Bayesian hierarchical additive model is available via a freely available R package, BHAM at (https://github.com/boyiguo1/BHAM)[https://github.com/boyiguo1/BHAM].

>>>>

Many applications of modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see @Mallick2013. These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients: @Ravikumar2009 extended the grouped lasso [@Yuan2006] to additive models; @Huang2010 further developed adaptive grouped lasso for additive models; @Wang2007 and @Xue2009 applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed where the application on AMs are by-products. @Scheipl2012 proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma, specifically for semi-parametric regression using MCMC. The normal-mixture-of-inverse gamma prior is mixture of t-distribution with an additional coefficients mixing vector, where the vector follows a mixture of two Gaussian distribution. @Bai2020 was the first to apply group spike-and-slab lasso prior to AMs using a fast optimization algorithm, and further generalized the framework to GAMs in @Bai2021. Regardless of the nice statistical properties, a practical and pivotal question, if non-linear effects exist, yet to be addressed by these methods. Besides @Schipel2012, most proposed methods adapt an 'all-in-all-out' strategy, i.e. either including or excluding the variable completely, rendering no space for within group selection. In addition, a common criticism received by these methods is that the basis function coefficients tend to be overly penalized due to the sparsity assumption directly applied on the basis function coefficients.[@Scheipl2013] It would be of special interest to develop a fast yet flexible generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we impose a new bi-part spike-and-slab spline prior on the smooth functions to perform bilevel selection, where the linear and nonlinear spaces of smooth functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. Additionally, two fast computing optimization based algorithms are developed for high and ultra-high dimension settings respectively. Meanwhile, to avoid the over-penalization of the basis function coefficients, smoothness penalties are incorporated in the model via re-parameterization. Smoothness penalties are commonly implemented in GAMs through smoothing and penalized regression splines. They are quadratic norm of the coefficients, and allow locally adaptive penalties on each smoothing function. For the most common case of cubic splines, the smoothness penalty conditioning on a smoothness penalty parameter $\lambda_j$ is a function of the integration of second derivative of the spline function, expressed mathematically as
$$
  \text{pen}f_j(x) = \lambda_j \int f^{"}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j,
$$
where $\bs S_j$ is a known penalty matrix, and $\bs \beta_j$ are the basis function coefficients. Smoothness penalties are also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothness penalty allows the separation of the linear space of the smooth function, also know as the null space, from the nonlinear space. @Marra2011 introduced a re-parameterization of the basis function matrix based on the decomposition of the smoothness penalty matrix $\bs S$. The re-parameterization allows different penalties imposed on the null and nonlinear spaces separately, and further allows the sparsity in linear space and smoothness in nonlinear effect interpolation. [TODO: a sentence that it decompose the prior to independent priors, provide obvious computational convenience.] The computational convenience improved the scalability of the model.

The proposed BHAM uses a spike-and-slab prior for function selections, but it substantially differs from previous spike-and-slab based GAMs, i.e. the spike-and-slab GAM [@Scheipl2012] and the SB GAM [@Bai2021] in many ways. First of all, the proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using independent mixture double exponential distribution, compared to spike-and-slab GAM. Spike-and-slab lasso priors provide computational convenience during model fitting through optimization algorithms, instead of computational intensive sampling algorithms. They make fitting high- and ultrahigh-dimensional model feasible without sacrificing performance in prediction and variable selection. Secondly, the optimization algorithms introduced in this article include EM-Coordinate Descent algorithm and EM-Iterative weighted least square algorithm for different utilities. The applications of the two aforementioned algorithm are fruitful in Bayesian hierarchical models in high-dimensional data analysis, see [TODO: add citations]. Even though SB-GAM with a group spike-and-slab lasso prior also uses EM-CD for model fitting, the proposed EM-IWLS provides uncertainty measures that could be further used for confidence bound calculation and hypothesis testing. EM-IWLS complements the lack of inference potential from EM-CD in a median and high data setting. Furthermore, the proposed model addresses the incapability of bilevel selection in SB-GAM. Lastly, the R implementation of the Bayesian hierarchical additive model is available publicly via BHAM at [https://github.com/boyiguo1/BHAM](https://github.com/boyiguo1/BHAM), make transnational science more accessible.

@@

<!-- Lastly, The third aim of the dissertation work will focus on the software implementation of the proposed models and algorithms. We will implement the proposed models and algorithms in one of the most commonly used statistical computing environment R.  The R implementation of the Bayesian hierarchical additive model is available via a freely available R package, BHAM at (https://github.com/boyiguo1/BHAM)[https://github.com/boyiguo1/BHAM].  -->

### 2021-06-23 Manuscript/SS_GAM.tex
More introduction writting

Many applications of modern biomedical research, e.g.~sequencing data
analysis, electric health record data analysis, require special
treatment of high-dimensionality, commonly known as \(p >> n\) problem.
There is extensive literature on high-dimensional linear models via
penalized models or Bayesian hierarchical models, see
\citet{Mallick2013}. These models assume sparsity among the predictors,
i.e.~most predictors have small or no effects on the outcome. Further
development focuses on addressing the grouping structure among the
predictors, where AMs and GAMs can be considered special cases. Various
penalties are imposed on the basis function coefficients:
\citet{Ravikumar2009} extended the grouped lasso \citep{Yuan2006} to
additive models; \citet{Huang2010} further developed adaptive grouped
lasso for additive models; \citet{Wang2007} and \citet{Xue2009} applied
grouped SCAD penalty \citep{Fan2001} to the additive models. Recent
Bayesian hierarchical models are also used in the context of
high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive MCMC algorithms
\citep{Xu2015, Yang2020} are proposed where the application on AMs are
by-products. \citet{Bai2020} was the first to apply group spike-and-slab
lasso prior to AMs using a fast optimization algorithm, and further
generalized the framework to GAMs in \citet{Bai2021}. Regardless of the
nice statistical properties, a practical and pivotal question, if
non-linear effects exist, yet to be addressed by these methods. In
addition, a common criticism received by these methods is that the basis
function coefficients tend to be overly penalized due to the sparsity
assumption directly applied on the basis function coefficients.
\citep{Scheipl2013}

To address this problem.

where majority of the efforts ware focusing on the

{[}TODO a sentence of sparsity. What it means in a penalized model and
Bayesian model{]} Directly apply to additive models, spline models
specifically. Problem, oversmoothing of the spline function. An more
appropriate treatment would be incorporating the smooth penalty of
spline functions.{[}TODO: a sentence about smoothness penalty{]}

What is the smooth penalty of spline function?

Many statistical models have been proposed for such problem, for
example, penalized models and Bayesian hierarchical models. However, up
to this point, most efforts of methodology development focus on modeling
linear effect in high dimensional setting. It is still not clear what is
the best practice on how to model non-linear effects in high-dimensional
setting.

Similar to expanding classic models to high-dimensional models,
expanding the GAM framework for high-dimensional applications is not a
trivial task. The difficulties mainly come in two-fold: grouping nature
of the basis functions, and the balance between spline smoothness and
variable sparsity. For GAM models, a smoothing function is consists of
multiple basis functions. The basis functions of a variable is
inherently related. The grouping structure should be properly taken care
of for accurate estimation. Moreover, shrinkage is previously used
solely for sparsity of the model in high dimensional regression models,
or smoothness in GAMs. How to organically combine the two together
remains as a troubling question.

To the best of our knowledge, there are few GAM models proposed for high
dimensional application. In common practice, model matrix of spline
terms are generated using base functions of preference, and then feed to
a penalized model. Without proper handle of the grouping nature of bases
functions, the estimation of the smooth function is less accurate.
\citet{Bai2020Spline} shows inferior performance of such practice
compared to using an ad hoc high-dimensional additive model. Among the
specialized high-dimensional spline methods, \citet{Ravikumar2009}
extended the original backfitting algorithm \citep{Hastie1987} to
accommodate the high-dimensional setting. The sparsity and smoothness
was handled separately with a soft threshold. The authors also
demonstrated its similarity to Group Lasso. \citet{Meier2009} proposed
an a sparsity-smoothness penalty,

\[
J(f_j) = \lambda_1 \sqrt{||f_j||^2_n + \lambda_2 \int(f^{''}_j(x))^2dx},
\] which can re-formulated as a group lasso problem as well. However,
the model can not estimate the smoothness of the spline functions
correctly, as the smoothness of all spline models are controlled by the
lambda\_2. The inflexible of the model creates problems in prediction.

From the Bayesian perspective, recent works that expands to the
high-dimensional settings includes \citet{Scheipl2012} and
\citet{Bai2020Spline}. \citet{Scheipl2012} proposed an new
spike-and-slab structure prior, normal-mixture-of-inverse gamma for
semi-parametric regression using MCMC. The normal-mixture-of-inverse
gamma prior is mixture of t-distribution with an additional coefficients
mixing vector, where the vector follows a mixture of two Gaussian
distribution. \citet{Bai2020Spline} proposed an novel high dimensional
Bayesian GAM using the spike-and-slab group lasso prior. An fast fitting
EM-based algorithm is proposed. Asymptotic theory, such as posterior
contraction rates and model selection consistency are proved. Even
though, the algorithm is flexible to cover many non-Gaussian family in a
high dimensional setting, the proposed method didn't address uncertainty
quantification, i.e.~calculating credible intervals etc. However, most
of these Bayesian GAMs still suffer a common problem, computation cost.
These Bayesian GAMs uses MCMC to estimate the posterior distribution,
which would create an intense computation load.

In this article, we aim to address the challenges of non-linear signal
modeling in high-dimensional application using Bayesian Hierarchical
model. Specifically, we will propose a new Bayesian spike-and-slab prior
for the generalized additive models and Cox spline models. The
contribution of the proposed prior comes in two-fold: 1) it supports
simultaneous variable selection and function selection; 2) it provides
robust estimation and prediction for the coefficients and outcomes. As
previously described, simultaneous variable selection and its functional
form selection is not a trivial problem. Nevertheless, the Bayesian
spike-and-slab prior is a perfect solution to the problem, as the latent
binary indicator can be used to for variable selection and the spike and
slab densities are used to adaptive adjust the smoothness of the spline
function. Moreover, compared to currently available Bayesian spline
models which mostly conduct a `all-in-all-out' strategy, i.e.~either
including a non-linear form of a variable or excluding the variable
completely, we offer a more flexible solution, rending one of three
possibilities: no effect of a predictor, only linear effect of a
predictor, or non-linear effect. This simplifies the process of
determining what effect a predictor have on the outcome, without
involving hypothesis testing where the degree of freedom for non-linear
effect is complicated to estimate due to the shrinkage. Moreover, out
model inherent the the robust estimation from Bayesian models. Lastly,
The third aim of the dissertation work will focus on the software
implementation of the proposed models and algorithms. We will implement
the proposed models and algorithms in one of the most commonly used
statistical computing environment R. Ancillary functions will be
provided in assistance to model diagnostic, interpretation and graphing.
The R implementation of the Bayesian hierarchical additive model is
available via a freely available R package, BHAM at
(\url{https://github.com/boyiguo1/BHAM}){[}\url{https://github.com/boyiguo1/BHAM}{]}.

The rest of the paper are arranged as the followings: in Chapter 2 we
demonstrate the Bayesian Hierarchical generalized additive model, where
we provide the setup of the proposed model and two fast-fitting
algorithms. In Chapter 3, we Application to real-world datasets
demonstrates the usefulness of the proposed methodologies in Chpater 4.
Conclusion and discussions are made in Chapter 5.

>>>>

Many applications of modern biomedical research, e.g.~sequencing data
analysis, electric health record data analysis, require special
treatment of high-dimensionality, commonly known as \(p >> n\) problem.
There is extensive literature on high-dimensional linear models via
penalized models or Bayesian hierarchical models, see
\citet{Mallick2013}. These models assume sparsity among the predictors,
i.e.~most predictors have small or no effects on the outcome. Further
development focuses on addressing the grouping structure among the
predictors, where AMs and GAMs can be considered special cases. Various
penalties are imposed on the basis function coefficients:
\citet{Ravikumar2009} extended the grouped lasso \citep{Yuan2006} to
additive models; \citet{Huang2010} further developed adaptive grouped
lasso for additive models; \citet{Wang2007} and \citet{Xue2009} applied
grouped SCAD penalty \citep{Fan2001} to the additive models. Recent
Bayesian hierarchical models are also used in the context of
high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed where the application
on AMs are by-products. \citet{Scheipl2012} proposed an ad hoc
spike-and-slab structure prior, normal-mixture-of-inverse gamma,
specifically for semi-parametric regression using MCMC. The
normal-mixture-of-inverse gamma prior is mixture of t-distribution with
an additional coefficients mixing vector, where the vector follows a
mixture of two Gaussian distribution. \citet{Bai2020} was the first to
apply group spike-and-slab lasso prior to AMs using a fast optimization
algorithm, and further generalized the framework to GAMs in
\citet{Bai2021}. Regardless of the nice statistical properties, a
practical and pivotal question, if non-linear effects exist, yet to be
addressed by these methods. Besides \citet{Schipel2012}, most proposed
methods adapt an `all-in-all-out' strategy, i.e.~either including or
excluding the variable completely, rendering no space for within group
selection. In addition, a common criticism received by these methods is
that the basis function coefficients tend to be overly penalized due to
the sparsity assumption directly applied on the basis function
coefficients.\citep{Scheipl2013} It would be of special interest to
develop a fast yet flexible generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we impose a new bi-part spike-and-slab spline prior on the
smooth functions to perform bilevel selection, where the linear and
nonlinear spaces of smooth functions can be selected separately. The
prior setup encourages a flexible solution, rending one of three
possibilities: no effect, only linear effect, or nonlinear effect of a
predictor. Additionally, two fast computing optimization based
algorithms are developed for high and ultra-high dimension settings
respectively. Meanwhile, to avoid the over-penalization of the basis
function coefficients, smoothness penalties are incorporated in the
model via re-parameterization. Smoothness penalties are commonly
implemented in GAMs through smoothing and penalized regression splines.
They are quadratic norm of the coefficients, and allow locally adaptive
penalties on each smoothing function. For the most common case of cubic
splines, the smoothness penalty conditioning on a smoothness penalty
parameter \(\lambda_j\) is a function of the integration of second
derivative of the spline function, expressed mathematically as \[
  \text{pen}f_j(x) = \lambda_j \int f^{"}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j,
\] where \(\boldsymbol{S}_j\) is a known penalty matrix, and
\(\boldsymbol{\beta}_j\) are the basis function coefficients. Smoothness
penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of the smooth function, also know as
the null space, from the nonlinear space. \citet{Marra2011} introduced a
re-parameterization of the basis function matrix based on the
decomposition of the smoothness penalty matrix \(\boldsymbol{S}\). The
re-parameterization allows different penalties imposed on the null and
nonlinear spaces separately, and further allows the sparsity in linear
space and smoothness in nonlinear effect interpolation. {[}TODO: a
sentence that it decompose the prior to independent priors, provide
obvious computational convenience.{]} The computational convenience
improved the scalability of the model.

The proposed BHAM uses a spike-and-slab prior for function selections,
but it substantially differs from previous spike-and-slab based GAMs,
i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting through optimization algorithms, instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional model feasible without sacrificing performance in
prediction and variable selection. Secondly, the optimization algorithms
introduced in this article include EM-Coordinate Descent algorithm and
EM-Iterative weighted least square algorithm for different utilities.
The applications of the two aforementioned algorithm are fruitful in
Bayesian hierarchical models in high-dimensional data analysis, see
{[}TODO: add citations{]}. Even though SB-GAM with a group
spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high data setting. Furthermore, the proposed model addresses the
incapability of bilevel selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

The rest of the paper are arranged as the followings: in Chapter 2 we
demonstrate the Bayesian hierarchical generalized additive model, where
we provide the setup of the proposed model and describe the two
fast-fitting algorithms. In Chapter 3, we compare the proposed via monte
carlo simulation. .Application to real-world datasets demonstrates the
usefulness of the proposed methodologies in Chpater 4. Conclusion and
discussions are made in Chapter 5.
